---
title: "Lab 13 - Colonizing Mars"
author: "Benjamin Egan"
date: "Start date 1-29"
output: github_document
---

```{r load-packages, message=FALSE, warning = FALSE, echo=FALSE}
library(tidyverse) 
library(ggplot2)
library(MASS)

#df_colonists <- subset(df_colonists, select = -c(resilience.1, agreeableness.1)) use if you need to remove extra columns

```

### Exercise 1 - Simulating colonists

```{r 1.1 - colonist attributes}


set.seed(123)

age <- rnorm(100, mean = 30, sd = 5)

df_colonists <- data.frame(id = 1:100, age)
                                      

```



```{r 1.1 - visualizing base colonists}

data.frame(df_colonists) %>%
ggplot(
 aes(x=age)
)+
  theme_bw()+
  geom_histogram(binwidth=2,color="black", fill="blue") +
  labs(title = "Histogram of Colonists Ages by Seed",
       x = "Age", y = "Frequency")

  
```

This is an age distribution graph. It created a roughly normal distribution that is centered at 30. It looks like changing the seed makes the distribution more normal. The 100 seed looks normally distributed and the 1000 seed even more so. Looks like it's due to random chance, as 999 seed doesn't have the nice bellcurve shape.


```{r 1.3 assigning roles}

role <- rep(c("engineer", "scientist", "medic"), 
             each = 34, 
             length.out = 100)

df_colonists <- df_colonists %>%
  mutate(role)

print(role)
```
Here I assigned each of them roles. I chose this method mostly out of pure randomness. But I like that it listed out each role together.

#### Health can be measured using the MARSGAR
```{r 1.4 heathy, message = FALSE}

set.seed(123)

df_colonists$marsgar <- runif(100, min = 40, max = 100)

df_colonists %>%
ggplot( aes(
    x = age,
    y = marsgar
  ) )+
  theme_bw()+
  geom_point()+
  geom_smooth(method=lm, color = "black",se = FALSE)+
  labs(
      x = "Age",
      y = "MARSGAR",
      title = "Scatterplot of Age and Health"
  )

```


### Exercise 2 - Growing the colonists' skills

```{r 2.1 technical skills and visual, warning = FALSE, message = FALSE}

set.seed(1235)

df_colonists$technical_skills <- 2 * df_colonists$age + rnorm(100, mean = 0, sd = 1)


df_colonists %>%
lm(technical_skills = 2 * age + noise)

df_colonists %>%
  ggplot(aes(
    x = age,
    y = technical_skills
  ))+
  geom_point(alpha = .6)+
  geom_line(stat = "smooth", method="lm", alpha = .6, color = "blue", se = FALSE)+
  labs(
    x = "Age",
    y = "Technical Skill",
    title = "Scatterplot of Colonist Age and Technical Skills"
  )

```

Here we can see that age is about evenly distributed among the roles

```{r Plot age and tehcnical skill by role}

df_colonists %>%
  ggplot(aes(
    x = age,
    y = technical_skills,
    color = as.factor(role)
  ))+
  geom_point(alpha = .6)+
  scale_color_manual(values=c("blue", "red","green4"))+
  geom_line(stat = "smooth", method="lm", alpha = .6, color = "black", se = FALSE)+
  labs(
    x = "Age",
    y = "Technical Skill",
    color = "Role",
    title = "Scatterplot of Colonist Age and Technical Skills by Role"
  )


```


### Problem solving for each colonist

```{r problem solving}


d_engineer <- rnorm(100, mean = 40, sd = 5)
d_scientist <- rnorm(100, mean = 60, sd = 7)
d_medic <- rnorm(100, mean = 90, sd = 2)



df_colonists <- df_colonists %>%
mutate(problem_solving = case_when(
    role == "engineer" ~ d_engineer,
    role == "scientist" ~ d_scientist,
    role == "medic" ~ d_medic
  ))


df_colonists %>%
  ggplot(aes(
    x = problem_solving,
    color = role,
    fill = role
  ))+
scale_fill_manual(values=c("blue", "red","green4"))+
  scale_color_manual(values=c("blue", "red","green4"))+
  geom_density(alpha = 0.7) +
  theme_bw()+
  labs(
    y = NULL,
    x = "Level of Problem Solving"
  )
```

## Using mvrnorm

```{r simulate resilience and agreeableness, message = FALSE}

mean_traits <- c(50, 50)
cov_matrix <- matrix(c(100, 50, 
                       50, 100), ncol = 2)



traits_data <- mvrnorm(n = 100, mu = mean_traits, Sigma = cov_matrix, empirical = FALSE)

colnames(traits_data) <- c("resilience", "agreeableness")

df_colonists <- cbind(df_colonists, traits_data)


df_colonists %>%
  ggplot(aes(
    x = resilience,
    y = agreeableness
  ))+
    geom_point()+
  geom_line(stat = "smooth", method="lm", alpha = .6, color = "blue", se = FALSE)
  

```

I used the code provided to create a correlation matrix for the big 5

```{r big 5 matrix}

seed <- 123
set.seed(123)


n_colonists <- 100


var_names <- c( "EX", "ES", "AG", "CO", "OP")

mean_traits <- c(-.5, .5, .25, .5, 0)
sd_traits <- c(1, .9, 1, 1, 1)


cor_matrix_bigfive <- matrix(c(
1.0000, 0.2599, 0.1972, 0.1860, 0.2949,
0.2599, 1.0000, 0.1576, 0.2306, 0.0720,
0.1972, 0.1576, 1.0000, 0.2866, 0.1951,
0.1860, 0.2306, 0.2866, 1.0000, 0.1574,
0.2949, 0.0720, 0.1951, 0.1574, 1.0000
), nrow=5, ncol=5, byrow=TRUE,
dimnames=list(c("EX", "ES", "AG", "CO", "OP"), 
c("EX", "ES", "AG", "CO", "OP")))


eigen_values <- eigen(cor_matrix_bigfive)$values
is_positive_definite <- all(eigen_values > 0)


cov_matrix_bigfive <- cor_matrix_bigfive * (sd_traits %*% t(sd_traits))


simulated_data <- mvrnorm(n = 100, mu = mean_traits, Sigma = cov_matrix_bigfive)


simulated_data <- cbind.data.frame(colonist_id = 1:n_colonists,
seed = seed,
simulated_data)

simulated_data$seed <- NULL

view(simulated_data)
```


## Preparing for the unexpected

I tried using a while() loop, but chat suggested using a for() loop.
```{r simulating colonies of big five}

max_colonies <- 100
n_colonists <- 100
MARSid_counter <- 1

var_names <- c( "EX", "ES", "AG", "CO", "OP")

mean_traits <- c(-.5, .5, .25, .5, 0)
sd_traits <- c(1, .9, 1, 1, 1)

cor_matrix_bigfive <- matrix(c(
1.0000, 0.2599, 0.1972, 0.1860, 0.2949,
0.2599, 1.0000, 0.1576, 0.2306, 0.0720,
0.1972, 0.1576, 1.0000, 0.2866, 0.1951,
0.1860, 0.2306, 0.2866, 1.0000, 0.1574,
0.2949, 0.0720, 0.1951, 0.1574, 1.0000
), nrow=5, ncol=5, byrow=TRUE,
dimnames=list(c("EX", "ES", "AG", "CO", "OP"), 
c("EX", "ES", "AG", "CO", "OP")))


cov_matrix_bigfive <- cor_matrix_bigfive * (sd_traits %*% t(sd_traits))

all_colonies <- list()

for (colony in 1:max_colonies) {
  set.seed(123 + colony)
  
  traits_data <- mvrnorm(n = n_colonists, mu = mean_traits, Sigma = cov_matrix_bigfive)
  
  df <- data.frame(
    colony = colony,
    id_in_colony = 1:n_colonists,
    MARSid = MARSid_counter:(MARSid_counter + n_colonists - 1),
    seed = 123 + colony,
    traits_data
  )
  names(df)[5:9] <- var_names
  
  all_colonies[[colony]] <- df
  MARSid_counter <- MARSid_counter + n_colonists
}

big_simulated_data <- do.call(rbind, all_colonies)
big_simulated_data$seed <- NULL



```

```{r extracting means and sd}

colonies_means <- big_simulated_data %>% 
  group_by(colony) %>% 
  summarise(mean_ex = mean(EX),
            mean_es = mean(ES),
            mean_ag = mean(AG),
            mean_co = mean(CO),
            mean_op = mean(OP)
            )

means_long <- colonies_means %>%
  pivot_longer(cols = -colony)

colonies_means <- means_long %>%
  mutate(big_5 = case_when(
    name == "mean_ex" ~ "Extraversion",
    name == "mean_es" ~ "Neuroticism",
    name == "mean_ag" ~ "Agreeableness",
    name == "mean_co" ~ "Conscientiousness",
    name == "mean_op" ~ "Openess"
  ))

names(colonies_means)[names(colonies_means) == "value"] <- "Mean"

colonies_sd <- big_simulated_data %>% 
  group_by(colony) %>% 
  summarise(
            sd_ex = sd(EX),
            sd_es = sd(ES),
            sd_ag = sd(AG),
            sd_co = sd(CO),
            sd_op = sd(OP)
            )

sd_long <- colonies_sd %>%
  pivot_longer(cols = -colony)

colonies_sd <- sd_long %>%
  mutate(big_5 = case_when(
    name == "sd_ex" ~ "Extraversion",
    name == "sd_es" ~ "Neuroticism",
    name == "sd_ag" ~ "Agreeableness",
    name == "sd_co" ~ "Conscientiousness",
    name == "sd_op" ~ "Openess"
  ))

names(colonies_sd)[names(colonies_sd) == "value"] <- "Standard Deviation"



colonies_stats <- full_join(colonies_means, colonies_sd, by = c("colony", "big_5"))



view(colonies_stats)

```

```{r test}
means_big_data1 <- means_big_data %>%
  mutate(big_5 = case_when(
    c("Extraversion Mean","Extraversion SD") == TRUE ~ "Extraversion",
    c("Neuroticism Mean", "Neuroticism SD") == TRUE ~ "Neuroticism",
    c("Agreeableness Mean", "Agreeableness SD") == TRUE ~ "Agreeableness",
    c("Conscientiousness Mean", "Conscientiousness SD") == TRUE ~ "Conscientiousness",
    c("Openess Mean","Openess SD") == TRUE ~ "Conscinetousness"
  ))


```

```{r plotting means and sd}

# Plot: Mean of EX
p1 <- ggplot(means_big_data, aes(x = mean_ex)) +
  geom_histogram(color = "black", fill = "skyblue", bins = 20) +
  geom_vline(xintercept = -0.5, color = "red", linetype = "dashed") +
  annotate("text", x = -0.5, y = Inf, label = "Pop Mean = -0.5", vjust = 2, color = "red") +
  labs(title = "Distribution of Colony Mean Extraversion", x = "Mean EX", y = "Count")

# Plot: SD of EX
p2 <- ggplot(means_big_data, aes(x = sd_ex)) +
  geom_histogram(color = "black", fill = "orange", bins = 20) +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  annotate("text", x = 1, y = Inf, label = "Pop SD = 1", vjust = 2, color = "red") +
  labs(title = "Distribution of Colony SD of Extraversion", x = "SD EX", y = "Count")

# Plot: Correlation between EX and OP
p3 <- ggplot(means_big_data, aes(x = cor_ex_op)) +
  geom_histogram(color = "black", fill = "plum", bins = 20) +
  geom_vline(xintercept = 0.2949, color = "red", linetype = "dashed") +
  annotate("text", x = 0.2949, y = Inf, label = "Pop r = 0.295", vjust = 2, color = "red") +
  labs(title = "Distribution of r(EX, OP) Across Colonies", x = "Correlation", y = "Count")

# Show plots
p1
p2
p3

```



